Capstone Movielens Report

### 2025-05-26 reports/capstone-movielens.site/02.5.UME-model.Rmd
RMD report:  UMGYE Model completed.

We will implement the _Regularization_ method on our  models (starting from the current model) in two steps:

@@

We are going to use the [tune.model_param](#func.tune.model_param) function described in the [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report, passing the [regularize.test_lambda.UM_effect.cv](#func.regularize.test_lambda.UM_effect.cv) function as the value of the [fn_tune.test.param_value](#func.tune.model_param.params.fn_tune.test.param_value) parameter:

```{r, eval=FALSE}
lambdas <- seq(0, 1, 0.1)

@@

Here we are going to use the [model.tune.param_range](#func.model.tune.param_range) function described in the [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report, passing the [regularize.test_lambda.UM_effect.cv](#func.regularize.test_lambda.UM_effect.cv) function as the value of the [fn_tune.test.param_value](#func.model.tune.param_range.params.fn_tune.test.param_value) parameter:
```{r, eval=FALSE}
endpoints <- 
  get_fine_tune.param.endpoints(cv.UME.preset.result$tuned.result)

>>>>

We will implement the _Regularization_ method on our  models (starting from the current model) in following three steps:

@@

We are going to use the [tune.model_param](#func.tune.model_param) function described in the [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report, passing the [regularize.test_lambda.UM_effect.cv](#func.regularize.test_lambda.UM_effect.cv) function as the value of the [fn_tune.test.param_value](#func.tune.model_param.params.fn_tune.test.param_value) parameter.

Below we provide the most significant part of the code that performs this operation:
```{r, eval=FALSE}
lambdas <- seq(0, 1, 0.1)

@@

Here we are going to use the [model.tune.param_range](#func.model.tune.param_range) function described in the [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report, passing the [regularize.test_lambda.UM_effect.cv](#func.regularize.test_lambda.UM_effect.cv) function as the value of the [fn_tune.test.param_value](#func.model.tune.param_range.params.fn_tune.test.param_value) parameter.

Below we provide the most significant part of the code that performs this operation:
```{r, eval=FALSE}
endpoints <- 
  get_fine_tune.param.endpoints(cv.UME.preset.result$tuned.result)

### 2025-05-26 reports/capstone-movielens.site/02.7.UMGYE-model.Rmd
Minor fix for the UMGYE Model Building section.
RMD report:  UMGYE Model completed.

If we define $\gamma(r)$ as the _year effect_ for the year (here denotes by $r$) of rating a movie $j$ by a user $i$, the formula \@ref(eq:UMGE-model) describing the _UMGE Model_, for the current model, takes the following form:

@@

::: {.noteblock data-latex=""}
Some of the functions described in this section accepts the `lambda` parameter, which we will need later for the _Model Regularization_ method. We use the [mean_reg](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L63) function call as well, which we will also need for the *Regularization* techniques to apply to our models (for details, see the  [mean_reg](#func.mean_reg) function description in the Section [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report).
We have already explained that in the [UME Model Regularization] section. 
For now, we omit the `lambda` parameter, accepting its default value `lambda = 0`. Let's recall that in this case, the [mean_reg](#func.mean_reg) function is equivalent to the standard R function [base::mean](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/mean).
:::

@@

We have already explained the idea of the _Linear Model Regularization_ in the [UME Model Regularization] section above. Let's extend the concept outlined there to our current model.

In this case, the formula \@ref(eq:ME-penalty) for adding a penalty takes the following form:

\begin{equation}
\sum_{i,j} \left(y_{u,i} - \mu - \alpha_i - \beta_j - ...\right)^2 + \lambda \sum_{j} \beta_j^2 
(\#eq:YE-penalty)
\end{equation}

And the formula \@ref(eq:ME-regularized) for calculating the values of the _treatment effect_ that minimize the equation takes the following form:

\begin{equation}
\hat{\beta}_j(\lambda) = \frac{1}{\lambda + n_j} \sum_{i=1}^{n_i} \left(Y_{i,j} - \mu - \alpha_i\right)
(\#eq:YE-regularized)
\end{equation}

where $n_y$ is the number of ratings made in year $y$.

As stated in the [UME Model Regularization] section, we will implement the _Regularization_ method on our  current model in three steps:

@@

::: {.noteblock data-latex=""}
The [regularize.test_lambda.UMGY_effect.cv]() function described below are defined in the [Regularization]() section of the [UM-effect.functions.R]() script.
:::

##### [regularize.test_lambda.UMGY_effect.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UMGY-effect.functions.R#L96) Function {#func.regularize.test_lambda.UMGY_effect.cv}
\

This function calculates _RMSE_ of the _UMGYE Model_ using _5-Fold Cross Validation_ method for the given $\lambda$ parameter value:
```{r, eval=FALSE}
```

@@

We are going to use the [tune.model_param](#func.tune.model_param) function described in the [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report, passing the [regularize.test_lambda.UMGY_effect.cv](#func.regularize.test_lambda.UMGY_effect.cv) function as the value of the [fn_tune.test.param_value](#func.tune.model_param.params.fn_tune.test.param_value) parameter:

```{r, eval=FALSE}
```
```{r}
str(cv.UMGYE.preset.result)
cv.UMGYE.preset.result$best_result
```

::: {.noteblock data-latex=""}
The complete version of the source code provided in this section can be found in the [UMGYE Model Regularization: Pre-configuration](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script.
:::

Now, let's visualize the results of the $\lambda$ range preconfiguration:
``` {r}
# title = TeX(r'[UMGE Model Regularization: $\lambda$ Range Pre-configuration]')
```

@@

@@

Here we are going to use the [model.tune.param_range](#func.model.tune.param_range) function described in the [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report, passing the [regularize.test_lambda.UMGY_effect.cv](#func.regularize.test_lambda.UMGY_effect.cv) function as the value of the [fn_tune.test.param_value](#func.model.tune.param_range.params.fn_tune.test.param_value) parameter:
```{r, eval=FALSE}
```
```{r}
str(UMGYE.rglr.fine_tune.results)
UMGYE.rglr.fine_tune.results$best_result
```

::: {.noteblock data-latex=""}
The complete version of the source code provided in this section can be also found in the [Fine-tuning Step of the Regularization Method for the User+Movie Model](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L867) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::

Let's visualize the fine-tuning results:
```{r}
```
\newpage

@@

Now, we can calculate the _Regularized UMGYE Effect_ by retraining our model on the entire `edx` dataset with the best value of the $\lambda$ parameter we just calculated, for the definitive _Root Mean Squared Error_ calculation and use in subsequent models.
```{r, eval=FALSE}
```

::: {.noteblock data-latex=""}
The complete version of the source code provided in this section are available in the [Re-training Regularized UMGYE Effect Model for the best $\lambda$](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L910) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L910) script on _GitHub_.
:::

We calculate the _Root Mean Squared Error_ for the ultimately computed _UMGYE Effect_ using [calc_UMGY_effect_RMSE.cv](#func.calc_UMGY_effect_RMSE.cv) function described above as follows:
```{r, eval=FALSE}
```

Finally, we add the definitive result for the current model to our _Result Table_:
```{r, eval=FALSE}
RMSEs.ResultTibble.rglr.UMGYE <- RMSEs.ResultTibble.UMGYE |> 
  RMSEs.AddRow("Regularized UMGYE Effect Model", 
               UMGYE.rglr.retrain.RMSE,
               comment = "Computed for `lambda` = %1" |>
                 msg.glue(UMGYE.rglr.best_lambda))
```
```{r}
RMSE_kable(RMSEs.ResultTibble.rglr.UMGYE)
```

>>>>

If we define $\gamma(r)$ as the _year effect_ for the year (here denotes by $r$) of rating a movie $j$ by a user $i$, the formula \@ref(eq:UMGE-model) describing the _UMGE Model_, for the current model, takes the form:

@@

::: {.noteblock data-latex=""}
Some of the functions described in this section accepts the `lambda` parameter, which we will need later for the _Model Regularization_ method. We use the [mean_reg](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L63) function call as well, which we will also need for the *Regularization* techniques to apply to our models (for details, see the  [mean_reg](#func.mean_reg) function description in the [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report).
We have already explained that in the [UME Model Regularization] section. 
For now, we omit the `lambda` parameter, accepting its default value `lambda = 0`. Let's recall that in this case, the [mean_reg](#func.mean_reg) function is equivalent to the standard R function [base::mean](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/mean).
:::

@@

We have already explained the idea of _Linear Model Regularization_ in the [UME Model Regularization] section above. We have also seen how the formula \@ref(eq:ME-penalty) for adding a penalty to the _UME Model_ is transformed into the formula \@ref(eq:GE-penalty) for the _UMGE Model_. For the current model, this formula takes the form:

\begin{equation}
\sum_{i,j} \left(y_{u,i} - \mu - \alpha_i - \beta_j - g_{i,j} - \gamma(r)\right)^2 + \lambda \sum_{r} \gamma(r)^2
(\#eq:YE-penalty)
\end{equation}

And the formula \@ref(eq:ME-regularized) for calculating the values of the _treatment effect_ that minimizes the equation will take the form:

\begin{equation}
\hat{\gamma}(r, \lambda) = \frac{1}{\lambda + n_r} \sum_{i=1}^{n_{r}} \left(Y_{i,j} - \mu - \alpha_i - \beta_j - g_{i,j}\right)
(\#eq:YE-regularized)
\end{equation}

where $n_r$ is the number of ratings made in year $r$.

As with the previous models, we implement the _Regularization_ method for the current model in the following three steps:

@@

::: {.noteblock data-latex=""}
The [regularize.test_lambda.UMGY_effect.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UMGY-effect.functions.R#L128) function described below are defined in the [Regularization](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UMGY-effect.functions.R#L127) section of the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UMGY-effect.functions.R#L128) script.
:::

##### [regularize.test_lambda.UMGY_effect.cv](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UMGY-effect.functions.R#128) Function {#func.regularize.test_lambda.UMGY_effect.cv}
\

This function calculates _RMSE_ of the _UMGYE Model_ using _5-Fold Cross Validation_ for the given $\lambda$ parameter value:
```{r, eval=FALSE}
regularize.test_lambda.UMGY_effect.cv <- function(lambda){
  if (is.na(lambda)) {
    stop("Function: regularize.test_lambda.UMGY_effect.cv
`lambda` is `NA`")
  }

  train_UMGY_effect.cv(lambda) |>
    calc_UMGY_effect_RMSE.cv()
}
```

@@

We are going to use the [tune.model_param](#func.tune.model_param) function described in the [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report, passing the [regularize.test_lambda.UMGY_effect.cv](#func.regularize.test_lambda.UMGY_effect.cv) function as the value of the [fn_tune.test.param_value](#func.tune.model_param.params.fn_tune.test.param_value) parameter.

Below we provide the most significant part of the code that performs this operation:
```{r, eval=FALSE}
  lambdas <- seq(0, 512, 32)
  cv.UMGYE.preset.result <- 
    tune.model_param(lambdas, 
                     regularize.test_lambda.UMGY_effect.cv,
                     steps.beyond_min = 16)
```
```{r}
str(cv.UMGYE.preset.result)
cv.UMGYE.preset.result$best_result
```

::: {.noteblock data-latex=""}
The complete version of the source code provided in this section can be found in the [UMGYE Model Regularization: Pre-configuration](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L1493) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L1493) script.
:::

Now, let's visualize the results of the $\lambda$ range preconfiguration:
``` {r}
cv.UMGYE.preset.result$tuned.result |>
  data.plot(title = TeX(r'[UMGYE Model Regularization: $\lambda$ Range Pre-configuration]'),
              xname = "parameter.value", 
              yname = "RMSE", 
              xlabel = TeX(r'[$\lambda$]'), 
              ylabel = "RMSE")
```

@@

::: {.noteblock data-latex=""}
The complete version of the source code provided in this section can be found in the [Fine-tuning Step of the Regularization Method for the User+Movie Model](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L867) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R) script on _GitHub_.
:::

@@

Here we are going to use the [model.tune.param_range](#func.model.tune.param_range) function described in the [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report, passing the [regularize.test_lambda.UMGY_effect.cv](#func.regularize.test_lambda.UMGY_effect.cv) function as the value of the [fn_tune.test.param_value](#func.model.tune.param_range.params.fn_tune.test.param_value) parameter.

Below we provide the most significant part of the code that performs this operation:
```{r, eval=FALSE}
endpoints <- 
  get_fine_tune.param.endpoints(cv.UMGYE.preset.result$tuned.result)

UMGYE.loop_starter <- c(endpoints["start"], 
                        endpoints["end"], 
                        8)
UMGYE.loop_starter
#> [1]

cache.base_name <- "UMGYE.rglr.fine-tuning"

UMGYE.rglr.fine_tune.results <- 
  model.tune.param_range(UMGYE.loop_starter,
                         UMGYE.rglr.fine_tune.cache.path,
                         cache.base_name,
                         regularize.test_lambda.UMGY_effect.cv)
```
```{r}
str(UMGYE.rglr.fine_tune.results)
UMGYE.rglr.fine_tune.results$best_result
```

::: {.noteblock data-latex=""}
Under the hood, this code block internally calls the helper function [get_fine_tune.param.endpoints](#func.get_fine_tune.param.endpoints) described in the [Regularization: Common Helper Functions](#appndx_a.rglr.common_helper.functions) section of the [Appendix] to this report to figure out the boundaries of the range of values from the nearest neighborhood of the value corresponding to the best `RMSE` in the data passed in as a parameter.
:::

Let's visualize the fine-tuning results:
```{r}
UMGYE.rglr.fine_tune.results$tuned.result |>
  data.plot(title = "UMGYE Model Regularization: Fine-tuned result",
              xname = "parameter.value",
              yname = "RMSE",
              xlabel = TeX(r'[$\lambda$]'),
              ylabel = str_glue("Deviation from the best RMSE value (",
                                as.character(round(UMGYE.rglr.fine_tune.RMSE.best, digits = 7)),
                                ")"),
              normalize = TRUE)
```

@@

::: {.noteblock data-latex=""}
The complete version of the source code provided in this section are available in the [Re-train Regularized UMGYE Model for the best `lambda`](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L1589) section of the [capstone-movielens.main.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/capstone-movielens.main.R#L1589) script on _GitHub_.
:::

Now, we can calculate the _Regularized UMGYE Effect_ by retraining our model on the entire `edx` dataset with the best value of the $\lambda$ parameter we just calculated, for the definitive _Root Mean Squared Error_ calculation and use in subsequent models.

```{r, eval=FALSE}
  rglr.UMGY_effect <- edx |> train_UMGY_effect(UMGYE.rglr.best_lambda)

::: {.noteblock data-latex=""}
Here we use the helper function [train_UMGY_effect](#func.train_UMGY_effect) described above in the [UMGYE Model Regularization: Support Functions] section.
:::

We calculate the _Root Mean Squared Error_ for the ultimately computed _UMGYE Effect_ using the helper function [calc_UMGY_effect_RMSE.cv](#func.calc_UMGY_effect_RMSE.cv) described above in the [UMGYE Model Regularization: Support Functions] section as follows:
```{r, eval=FALSE}
  rglr.UMGY_effect.RMSE <- calc_UMGY_effect_RMSE.cv(rglr.UMGY_effect)
```

Finally, we add the definitive result for the current model to our _Result Table_:
```{r, eval=FALSE}
RMSEs.ResultTibble.rglr.UMGYE <- RMSEs.ResultTibble.UMGYE |> 
  RMSEs.AddRow("Regularized UMGYE Model", 
               rglr.UMGY_effect.RMSE,
               comment = "Computed for `lambda` = %1" |>
                 msg.glue(UMGYE.rglr.best_lambda))

```
```{r}
RMSE_kable(RMSEs.ResultTibble.rglr.UMGYE)
```

### 2025-05-26 reports/capstone-movielens.site/04.appendix.Rmd
RMD report:  UMGYE Model completed.

Next, the function figures out the boundary indices of a range of values from the neighborhood of the minimum `RMSE` by calling internally another helper function `get_fine_tune.param.endpoints.idx` (see the description of the function [get_fine_tune.param.endpoints.idx](#func.get_fine_tune.param.endpoints.idx) below) and, using one more helper function `get_best_param.result`  (see the description of the function [get_best_param.result](#func.get_best_param.result) below), a pair of values,  corresponding to the best result: minimal  `RMSE` and corresponding input parameter value (which is considered the best).

@@

>>>>

Next, the function figures out the boundary indices of a range of values from the neighborhood of the minimum `RMSE` by calling internally another helper function [get_fine_tune.param.endpoints.idx](#func.get_fine_tune.param.endpoints.idx) (see the description below) and, using one more helper function [get_best_param.result](#func.get_best_param.result)  (see the description below), a pair of values,  corresponding to the best result: minimal  `RMSE` and corresponding input parameter value (which is considered the best).

@@

### [get_fine_tune.param.endpoints](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L80) Function {#func.get_fine_tune.param.endpoints}
\

The algorithm finds the values of the nearest neighboring elements of the element corresponding to the minimum RMSE value in the given sequence.

#### Signature
\
```{r eval=FALSE}
get_fine_tune.param.endpoints <- function(preset.result) {
  # ...

  c(start = preset.result$parameter.value[i], 
    end = preset.result$parameter.value[j],
    best = preset.result$parameter.value[best.idx])
}
```

::: {#func.get_fine_tune.param.endpoints.params.preset.result .sidebar }
##### ***preset.result***
\
A data structure compatible with the `tuned.result` item of the data structure returned by the [tune.model_param](#func.tune.model_param) function (see the [Value](#func.tune.model_param.value) subsection of the `tune.model_param` function description section for more details). Here is a sample of using this parameter in the `model.tune.param_range` function [code](#func.model.tune.param_range.code):
```{r eval=FALSE}
    tuned_result <- tune.model_param(test_param_vals, 
                                  fn_tune.test.param_value,
                                  break.if_min,
                                  steps.beyond_min)

    tuned.result <- tuned_result$tuned.result
    # tuned.result = data.frame(RMSE, parameter.value)
    # ...
    bound.idx <- get_fine_tune.param.endpoints(tuned.result)
    # bounnd.idx =  c(start.index, end.index, min_RMSE.index)
```
:::

#### Details
\

Under the hood, the function internally calls another helper function [get_fine_tune.param.endpoints.idx](#func.get_fine_tune.param.endpoints.idx) described below to figure out the boundary indices of the range of values from the nearest neighborhood of the value corresponding to the best `RMSE` in the data passed in the [preset.result](#func.get_fine_tune.param.endpoints.params.preset.result) parameter.

#### Value
\
A vector containing the element value corresponding to the minimum `RMSE` in a given sequence, along with the values of its nearest neighboring elements:
```{r eval=FALSE}
  c(start = preset.result$parameter.value[i], 
    end = preset.result$parameter.value[j],
    best = preset.result$parameter.value[best.idx])
```

#### Source Code
\
The source code of the [get_fine_tune.param.endpoints](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/common-helper.functions.R#L80) function is provided below:
```{r eval=FALSE}
get_fine_tune.param.endpoints <- function(preset.result) {

  preset.result.idx <- get_fine_tune.param.endpoints.idx(preset.result)

  i <- preset.result.idx["start"]
  j <- preset.result.idx["end"]
  best.idx <- preset.result.idx["best"]

  c(start = preset.result$parameter.value[i], 
    end = preset.result$parameter.value[j],
    best = preset.result$parameter.value[best.idx])
}
```
